### 2020-03-04 doc/icfp20/harmony.tex
tweak harmony section

The primary Example~9.1 in Piston's text is shown on the left
side of Figure~\ref{fig:harmonization} which shows the melody with a
sample four part SATB harmonization in C~major. The melody is the
series of highest notes, the soprano (S) line, while the other four
parts alto (A), tenor (T) and bass (B) are the melodies below
it. Roman numerals denote the names of the triads at each vertical
slice. Note that even though there are four voices there are only
three distinct pitches sounding at each vertical slice (where tones an
octave apart are considered to have the same relative pitch). As is
typical the root of the triad is always doubled. Also note that all
triads are in root position (meaning the root is in the bass voice),
as Piston restricts to this case initially in the chapter (he does
relax it later). The goal of a harmonization is both to create an
aesthetically pleasing chord progression as well as fluid melodies in
each voice (called voice leading) and independence of each music
line (in other words good counterpoint). In this chapter as well as
throughout the text Piston gives advice on how to achieve these goals.

@@

We would like to formalize some of this advice and apply it to
automate harmonization of melodies. We start with selection of the
sequence of triads, in other words the harmonic progression. Earlier
in the text on page~23 Piston presents a Table of Usual Root
Prrogressions, noting which triads usually follow a given one and
which are sometimes used or rare. It is easy to encode this in Agda.

\begin{alltt}
data Triad : Set where
  I : Triad; II : Triad; III : Triad; IV : Triad; V : Triad; VI : Triad; VII : Triad

record NextTriad : Set where
  constructor nextTriadb
  field
    usual : List Triad; sometimes : List Triad; rare : List Triad

@@

It is easiest to create a harmonic progression in reverse, to
ensure one ends a musical phrase with either a full (V--I) or half (V)
cadence, and work backward from there. So we would like to invert this
table to give a list of likely triads to precede a given triad. This
would normally be straightfoward but here we see a disadvantage of
using Agda. As is typical with languages based on dependent type
theories, equality is a very subtle concept and even determing if a
term is a member of a list is nontrivial.

In Haskell this would be easy--there is an \texttt{Eq} typeclass and a
method for automatically generating equalities. In the Agda Standard
Library~\citep{agda-stdlib} is a relation with constructive proof of
membership, and requires far more effort than we need. Alternatives
include defining decidable equality for triads (which requires writing
out 49 cases) or mapping to natural numbers and using already
implemented decidable or boolean equality for those. For now we simply
write the inverse function manually, excluding the rare cases.

@@

TODO: Finish this.

@@

Most of the the Haskell-based
work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
the problem of harmonization of a melody, and their harmonization of
Piston's example is shown on the right side of
Figure~\ref{fig:harmonization}. Their method is quite different from
ours. Instead of using Piston's table of root progressions they use a
more sophisticated hierarchical grammar from previous
work~\citep{magalhaes-harmtrace}.

Their technique is to first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to
create a harmonic analysis of every possible sequence of chords that
arises, choosing the one that most closely fits their
grammar. Although the result is not bad, it clearly leaves room for
improvement. Noted in the paper is that no attention is given to
voicing (how each of the four parts flows horizontally and how the
lines interact---in other words the counterpoint). Not noted is that
the melody note is always doubled, whereas typically in four-part
harmony it is preferred to double the root note. Also minimizing the
number of parse errors as a means to choose the best harmonization
does not necessarily have a musical meaning.

>>>>

The primary Example~9.1 in Piston's text
(Figure~\ref{fig:harmonization} left) shows a
sample four part SATB harmonization in C~major. The melody is the
series of highest notes, i.e., the soprano (S) line, while the other three
parts alto (A), tenor (T), and bass (B) are the melodies below
it. Roman numerals denote the names of the triads at each vertical
slice. Note that even though there are four voices there are only
three distinct pitches sounding at each vertical slice (where tones an
octave apart are considered to have the same relative pitch). As is
typical the root of the triad is always doubled. Also note that all
triads are in root position (meaning the root is in the bass voice),
as Piston restricts to this case initially in the chapter (he does
relax it later). The goal of a harmonization is both to create an
aesthetically pleasing chord progression as well as fluid melodies in
each voice (called voice leading) and independence of each music
line (in other words good counterpoint). In this chapter
Piston gives advice on how to achieve these goals.

@@

We would like to formalize some of this advice and apply it to
automate harmonization of melodies. We start with selection of the
sequence of triads, in other words the harmonic progression. Earlier
in the text (on page~23) Piston presents a Table of Usual Root
Progressions, noting which triads usually follow a given one and
which are sometimes used or rare. It is easy to encode this in Agda
using the \texttt{Triad} datatype from Section~\ref{sec:music}.

\begin{alltt}
record NextTriad : Set where
  constructor nextTriadb
  field
    usual : List Triad; sometimes : List Triad; rare : List Triad

@@

It is easiest to create a harmonic progression in reverse, to
ensure that one ends a musical phrase with either a full (V--I) or half (V)
cadence, and work backward from there. So we would like to invert this
table to give a list of likely triads to precede a given triad. This
would normally be straightfoward but here we see a disadvantage of
using Agda. As is typical with languages based on dependent type
theories, equality is a very subtle concept and even determing if a
term is a member of a list is nontrivial.

In Haskell this would be easy---there is an \texttt{Eq} typeclass and a
method for automatically generating equalities. In the Agda Standard
Library~\citep{agda-stdlib} is a relation with a constructive proof of
membership, and requires far more effort than we need. Alternatives
include defining decidable equality for triads (which requires writing
out 49 cases) or mapping to natural numbers and using already
implemented decidable or boolean equality for those. For now we simply
write the inverse function manually, excluding the rare cases.

@@

As Piston does we require the harmonizing triad to include the melody
note. We use a function \texttt{harmonizations} to generate a list of
all candidate harmonizations for a given melody. For space
considerations we refer the reader to the supplement
\texttt{Harmony.agda} for its definition. Here again we encounter difficulties
with equality and membership. We use a bit vector to represent the
set of scale degrees in a triad, where each scale degree is mapped to
an element of \texttt{Fin 7}.

We next restrict to progressions ending in a half cadence (V) to match
Piston's example. This results in 25 candidate progressions.

@@

Most of the Haskell-based
work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
the problem of harmonization of a melody;
Figure~\ref{fig:harmonization} right shows their harmonization of
Piston's example. Their method is quite different from
ours. Instead of using Piston's table of root progressions they use a
more sophisticated hierarchical grammar from previous
work~\citep{magalhaes-harmtrace}.
More specifically, they first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to create
a harmonic analysis of every possible sequence of chords that arises,
choosing the one that most closely fits their grammar.

Although the result produced by \fharm is not bad, it clearly leaves
room for improvement. Noted in
the paper is that no attention is given to voicing (how each of the
four parts flows horizontally and how the lines interact---in other
words the counterpoint). They use inversions (in which the
third or fifth of the chord is in the bass) to ensure each melody line
does not move too much, but as the harmonizing triads are all in close
postion separated from the melody, the result sounds like a soprano
melody with block chord accompaniment, rather than four independent
lines.

Furthermore, \fharm always double the melody note, not the root note
as Piston does (and as preferred in four-part harmony in general).
Also, minimizing the number of parse errors as a means
to choose the best harmonization does not necessarily have a musical
meaning.
